---
title: "Data Science 1 Homework 2"
author: "Samina Rashiq"
date: "2024-10-06"
output: github_document
---
  
```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(tidyr)
library(knitr)
setwd("/Users/saminarashiq/R/Data_Science_1")
```

# Problem 1
```{r nyctransit, results='hide'}
nyc_transit = read_csv("./NYC_Transit_Subway_Entrance_And_Exit_Data.csv", show_col_types = FALSE) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, 
    station_longitude, starts_with("route"), 
    entry, vending, entrance_type, ada
  ) |>
  mutate(
    entry_logical = ifelse(entry == "YES", TRUE, FALSE)
  )
```
This data set describes key details about the NYC subway system including station names, routes served, ADA compliance and more. I cleaned the data by using the janitor function to convert all names to lowercase and replacing all spaces/special characters with underscores. I then selected relevant variables and turned the entry variable from a character type to a logical type, leaving a cleaner dataset with 1,868 rows and 32 columns. 

```{r distinct_stations, echo=FALSE}
distinct_stations = nyc_transit |>
  distinct(station_name, line)

num_distinct_stations = nrow(distinct_stations)
cat("The number of distinct stations is:", num_distinct_stations)

num_ada_compliant = nyc_transit |>
  filter(ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
cat("The number of ADA-compliant stations is:", num_ada_compliant)

without_vending = nyc_transit |>
  filter(vending == "NO", entry_logical == TRUE) |>
  nrow()

total_without_vending = nyc_transit |>
  filter(vending == "NO") |>
  nrow()

proportion_without_vending_entry = without_vending / total_without_vending
cat("The proportion of station entrances/exits with no vending that allow entrance is:", proportion_without_vending_entry)

```

```{r a_stations, echo=FALSE}
a_stations = nyc_transit |>
  mutate(across(route1:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route"
  ) |> 
  filter(route == "A") |>
  select(station_name, line) |>
  distinct()

num_a_stations = nrow(a_stations)
cat("The number of distinct stations that serve the A train is:", num_a_stations)

ada_compliant_a_stations = a_stations |>
  filter(station_name %in% (nyc_transit |> filter(ada == TRUE) |> pull(station_name)))

num_ada_a_stations = nrow(ada_compliant_a_stations)
cat("The number of ADA-compliant stations that serve the A train is:", num_ada_a_stations)

```






