---
title: "Data Science 1 Homework 2"
author: "Samina Rashiq"
date: "2024-10-06"
output: github_document
---
```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(tidyr)
library(knitr)
library(dplyr)
library(janitor)
```

# Problem 1
```{r nyctransit, results='hide'}
nyc_transit = read_csv("./NYC_Transit_Subway_Entrance_And_Exit_Data.csv", show_col_types = FALSE) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, 
    station_longitude, starts_with("route"), 
    entry, vending, entrance_type, ada
  ) |>
  mutate(
    entry_logical = ifelse(entry == "YES", TRUE, FALSE)
  )
```
This data set describes key details about the NYC subway system including station names, routes served, ADA compliance and more. I cleaned the data by using the janitor function to convert all names to lowercase and replacing all spaces/special characters with underscores. I then selected relevant variables and turned the entry variable from a character type to a logical type, leaving a cleaner dataset with 1,868 rows and 32 columns. 

```{r distinct_stations, echo=FALSE}
distinct_stations = nyc_transit |>
  distinct(station_name, line)

num_distinct_stations = nrow(distinct_stations)
cat("The number of distinct stations is:", num_distinct_stations)

num_ada_compliant = nyc_transit |>
  filter(ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
cat("The number of ADA-compliant stations is:", num_ada_compliant)

without_vending = nyc_transit |>
  filter(vending == "NO", entry_logical == TRUE) |>
  nrow()

total_without_vending = nyc_transit |>
  filter(vending == "NO") |>
  nrow()

proportion_without_vending_entry = without_vending / total_without_vending
cat("The proportion of station entrances/exits with no vending that allow entrance is:", proportion_without_vending_entry)

```

```{r a_stations, echo=FALSE}
a_stations = nyc_transit |>
  mutate(across(route1:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route"
  ) |> 
  filter(route == "A") |>
  select(station_name, line) |>
  distinct()

num_a_stations = nrow(a_stations)
cat("The number of distinct stations that serve the A train is:", num_a_stations)

ada_compliant_a_stations = a_stations |>
  filter(station_name %in% (nyc_transit |> filter(ada == TRUE) |> pull(station_name)))

num_ada_a_stations = nrow(ada_compliant_a_stations)
cat("The number of ADA-compliant stations that serve the A train is:", num_ada_a_stations)

```

# Problem 2
```{r mrtrash}
mr_trash_df = 
  read_excel("202409 Trash Wheel Collection Data.xlsx", 
    sheet = "Mr. Trash Wheel") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  select(where(~ !all(is.na(.)))) |>
  mutate(
    sports_balls = as.integer(
      round(sports_balls, 0)),
    year = as.numeric(year),
    trash_wheel = "Mr. Trash Wheel")
```

```{r proftrash}
prof_trash_df = 
  read_excel("202409 Trash Wheel Collection Data.xlsx", 
    sheet = "Professor Trash Wheel") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(trash_wheel = "Professor Trash Wheel")
```

```{r gwyntrash}
gwyn_trash_df = 
  read_excel("202409 Trash Wheel Collection Data.xlsx", 
    sheet = "Gwynnda Trash Wheel") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(trash_wheel = "Gwynnda Trash Wheel")

```

```{r merge_datasets}
trash_df =
  bind_rows(mr_trash_df, prof_trash_df, gwyn_trash_df)
```

The data set contains data collected by three different trash wheels in Baltimore, MD. The dataset includes `r nrow(trash_df)` observations. Some key variables in this dataset are `year`, which represents the year of trash collection, `weight_tons`, which is the weight of trash collected in tons, and `cigarette_butts`, which indicates the number of cigarette butts collected.

The total weight of trash collected by Professor Trash Wheel is `r sum(trash_df$weight_tons[trash_df$trash_wheel == "Professor Trash Wheel"], na.rm = TRUE)` tons. Additionally, the total number of cigarette butts collected by Gwynnda Trash Wheel in June 2022 is `r sum(trash_df$cigarette_butts[trash_df$trash_wheel == "Gwynnda Trash Wheel" & trash_df$month == "June" & trash_df$year == 2022], na.rm = TRUE)`.

# Problem 3
```{r bakers}
bakers_df = 
  read.csv("bakers.csv") |>
  janitor::clean_names() |>
  mutate(
    baker = word(baker_name, 1),
    baker = if_else(baker == "Jo" | baker == '"Jo"', "Jo", baker)
    ) |>
  select(baker, series, baker_age, baker_occupation, hometown) |>
  arrange(series)
```

```{r bakes}
bakes_df = 
  read.csv("bakes.csv") |>
  janitor::clean_names() |>
  select(baker, series, episode, signature_bake, show_stopper) |>
  mutate(baker = if_else(baker == "Jo" | baker == '"Jo"', "Jo", baker))
```

```{r results}
results_df = read.csv("results.csv", skip = 2) |>
  janitor::clean_names() |>
  select(series, episode, technical, baker, result)
```
```{r merging_bakers_bakes}
merged_bakers_bakes = bakes_df |>
  full_join(bakers_df, by = c("baker", "series"))
```

```{r merging_all}
final_bakeoff_data_df = merged_bakers_bakes |>
  full_join(results_df, by = c("baker", "series", "episode")) |>
  mutate(
    signature_bake = if_else(is.na(signature_bake), "Not Available", signature_bake),
    show_stopper = if_else(is.na(show_stopper), "Not Available", show_stopper),
    result = if_else(is.na(result), "No Result", result)
  )
```
```{r export_final_bakeoff_data_df}
write_csv(final_bakeoff_data_df, "final_bakeoff_data.csv")
```
I started by replacing '"Jo"' in the bakers.csv and bakes.csv datasets with 'Jo' to reduce the chance of errors. Since only first names were available in the bakes.csv and results.csv datasets, I extracted first names from bakers.csv. In results.csv, I skipped the first two rows as they did not contain headers or data.

I merged the data in two steps because it was easier for me to understand. First, I merged bakers.csv and bakes.csv on baker and series to connect each baker to their bakes. Then, I merged the output with results.csv using baker, series, and episode because  performance results are tied to specific episodes (which are not present in the bakers.csv dataset).


```{r sbaker_table}
baker_summary_df = final_bakeoff_data_df |>
  filter(series >= 5 & series <= 10) |>  
  group_by(series, baker) |>              
  summarise(
    star_baker_count = sum(result == "STAR BAKER"),
    winner = if_else(any(result == "WINNER"), "Yes", "No"),
    .groups = "drop"
  ) |>
  filter(star_baker_count > 0 | winner == "Yes") |>
  arrange(series, desc(star_baker_count))
```

### Star Baker Counts and Winners (Seasons 5-10) with Counts > 0 unless Winner ###

```{r sbaker_table_display, echo=FALSE}
baker_summary_df = baker_summary_df |>
  mutate(
    baker = if_else(winner == "Yes", paste0("**", baker, "**"), baker)
  )
library(knitr)
baker_summary_df |> 
  kable(
    col.names = c("Series", "Baker", "Star Baker Count", "Winner")
  )
```
It was surprising that Richard in Series 5 did not win with an all-time high of five star baker awards. It is also surprising that in season 10, David with zero star baker awards won over Steph, who had 4. Otherwise, the results seem quite reasonable. 

```{r viewers, echo=FALSE, message=FALSE, results='hide'}
viewers_long_df = read.csv("viewers.csv") |>
  janitor::clean_names() |>
  pivot_longer(
    cols = starts_with("series"),
    names_to = "series",
    names_prefix = "series_",
    values_to = "viewers"
  ) |>
  mutate(
    series = as.numeric(series)
  ) |>
  drop_na(viewers) |>
  arrange(series)

average_viewership_season_1 = viewers_long_df |>
  filter(series == 1) |>
  summarise(average_viewers = mean(viewers, na.rm = TRUE)) |>
  pull(average_viewers)

average_viewership_season_5 = viewers_long_df |>
  filter(series == 5) |>
  summarise(average_viewers = mean(viewers, na.rm = TRUE)) |>
  pull(average_viewers)
```

The average viewership in Season 1 was `r round(average_viewership_season_1, 2)` million viewers.

The average viewership in Season 5 was `r round(average_viewership_season_5, 2)` million viewers.

### First 10 Rows of Tidied Viwership Dataset ###
```{r finaltable}
library(knitr)
viewers_long_df |> 
  head(10) |>           
  kable(col.names = c("Episode", "Series", "Viewers (millions)"))
```